---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This documents aims respond some issues about [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). The data is about personal movement using activity monitoring devices such as [Fibit](http://www.fitbit.com/), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband)  or [Jawbone Up](https://jawbone.com/intl/up). The device collects data at 5 minutes intervals through out the day - consists of two months of data from an anonymous individual collect during Octover and November, 2012.

The variables included in this dataset (17,568 observations) are:

- **steps**: number of steps taking in a 5 minutes interval (missing values are code as NA)

- **date**: the date on which the measurement was taken in YYYY-MM-DD format

- **interval**: identifier for 5 minutes interval in which measurement was taken


####Loading and preprocessing the data

To begin, we read the file and store in *data* variable:

```{r}
data<-read.csv('activity.csv')
```

We, also, need some packages:

```{r,result='hide'}
library(stringr)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(scales)
```

We need to format *data$interval* in order to see the time interval,
```{r}
data$interval<-str_pad(data$interval,width=4,pad=0)
data$interval<- paste(str_sub(data$interval,1,2),':',str_sub(data$interval,3,4),sep='')
data$interval<-as.factor(data$interval)
```

The resultant data is
```{r}
head(data,3)
```


####What is mean total number of steps taken per day?

We organize the data grouping by *date*,
```{r}
data2<-group_by(data,date)
```

The total number of steps taken each day is showed by the follow histogram

```{r, fig.height=4,fig.align='center'}
day_sum<-summarize(data2,s=sum(steps,na.rm=T))
head(day_sum)
hist(day_sum$s,breaks=20,main='Histogram of total number of steps per day',xlab='Steps')
```
Note we disconsider NA values.

The mean and median total number of steps taken per day are
```{r}
mean_totalstep<-mean(day_sum$s,na.rm=T)
mean_totalstep
median_totalstep<-median(day_sum$s,na.rm=T)
median_totalstep
```

####What is the average daily activity pattern?

We will analyze the average number of steps taken across the 5-minute interval, on a time series plot.

```{r,fig.align='center',fig.width=10,fig.height=4}
data3<-group_by(data,interval)
interval_mean<-summarize(data3,m=mean(steps,na.rm=T))
head(interval_mean)
plot(interval_mean$interval,interval_mean$m,type='l',ylab='steps',xlab='Interval',main='Daily activity pattern')
```

Note the 5-minute interval that contains the maximum number of steps can be founded by
```{r}
maximum<-max(interval_mean$m)
max_interval<-interval_mean$interval[interval_mean$m==maximum]
maximum
max_interval
```
The maximum value, `r maximum`, occurs at `r max_interval`.


####Imputing missing values

Some entrances are missing values (coded as `NA`). We want to verify the impact of missing values into the calculations; first, we calculate the total number of missing values

```{r}
l_na<- length(which(is.na(data)))
l_na
```

Around 13% of the data are missing values. To fill the missing values in the dataset, we replace by the mean dor 5-minute interval.

For start, we must attribute the dataset in another variable
```{r}
data_mod<-data
```
And replace the `NA` by the average of 5-minute interval,
```{r}
for(i in seq_along(data_mod$steps[which(is.na(data_mod))])){
    data_mod$steps[which(is.na(data_mod))]<-
        interval_mean$m[interval_mean$interval==data_mod$interval[which(is.na(data_mod))]]
}
head(data_mod,3)
```

Now we want verify the difference with the first histogram, organizing the data grouping by *date*,
```{r}
data2_mod<-group_by(data_mod,date)
```

The folow histogram shows the difference,
```{r, fig.height=4,fig.align='center'}
day_sum_mod<-summarize(data2_mod,s=sum(steps,na.rm=T))
head(day_sum_mod)
hist(day_sum_mod$s,breaks=20,main='Histogram of total number of steps per day',xlab='Steps')
```

Note the maximum frequency is around 1000, and in the last histogram, in 0 and 1000 with smaller frequency.

```{r}
mean_totalstep_mod<-mean(day_sum_mod$s,na.rm=T)
mean_totalstep_mod
median_totalstep_mod<-median(day_sum_mod$s,na.rm=T)
median_totalstep_mod
```
Another difference resides on the mean and on the median: `r mean_totalstep_mod` and `r median_totalstep_mod` *versus* `r mean_totalstep` and `r median_totalstep` founded before.

####Are there differences in activity patterns between weekdays and weekends?

To conclude, we want to see difference in the behavior in weekdays and weekends. We create a factor to represent what kind of day 
```{r,results='hide'}
days<-c(rep(c(rep('weekday',5),rep('weekend',2)),floor(length(data_mod$steps)/7)),rep('weekday',5))
factor(days,levels=c('weekday','weekend'))
```

```{r}
data_mod$days<-days
head(data_mod,7)
```

In order to analyze the difference between weekdays and weekends, we aggregate the data by type of days, as follows

```{r,fig.align='center',fig.width=10,fig.height=4}
steps_weekdays <- aggregate(steps ~ interval + days, data_mod, mean)
head(steps_weekdays)
ggplot(data = steps_weekdays,aes(as.POSIXct(interval,format="%H:%M"),steps)) + 
    geom_line() + facet_grid(days ~ .)  +
    scale_x_datetime(labels = date_format("%H:%M"),breaks = "4 hour")+
    xlab("Time [hour]") + ylab("Average for each hour")
```
We could see a smooth increase of activity in weekends in the afternoon, with peaks 16:00.
